---
title: 'Thesis: Exploratory Data Analysis'
author: "Laurens van der Tas"
output: html_document
---

### Initialize R Environment & Load Data

This section will run the commands necessary to initialize R and load up our packages and data.

```{r Initialize, results='hide', message=FALSE, warning=FALSE}
#Set working directory
setwd("C:/Users/Laurens/Dropbox/University/Year 4/Period 2/Applied Economics Research Course/Thesis/data")

#Load required packages
library(foreign)
library(stargazer)
library(ggplot2)
library(aod)
library(gridExtra)
library(ggthemes)
library(dplyr)
library(mfx)
library(corrplot)

#Enable anti-aliasing for rendered graphics
library(knitr)
#opts_chunk$set(out.width = '\\maxwidth')
#dev = "CairoPNG", 

#Load dataset
data.dropout <- read.dta("DatasetTrimmed.dta")
  #Read name vector of dataset
  names(data.dropout)
```

```{r Export Command, eval=FALSE, echo=FALSE}
#Regular ggplot export
ggsave(plot = zorg.1, 
       "output.png", 
       h = 9/3, 
       w = 16/3, 
       type = "cairo-png")

#R cairopng: run code below, run the plot, then run dev.off()
png(filename = "Rplot%03d.png",
    width = 1280, height = 1024, units = "px", pointsize = 12,
     bg = "white",  res = NA, 
    type = "cairo-png")

#R PDF export ggplot: usage same as above
pdf(file = "C:/Users/Laurens/Desktop/img.pdf",
    width = 12,
    height = 15,
    pointsize = 5)
```


### Descriptive Statistics

Here, the two dichotomous control variables are factorized.

```{r Descriptive 1}
#Factorize binaries
data.dropout$geslachtBin <- factor(data.dropout$geslachtBin, labels = c("Female", "Male"))
data.dropout$allochtoonBin <- factor(data.dropout$allochtoonBin, labels = c("No", "Yes"))
```

### Boxplots

The following section will create boxplots for the *Big Five* and *Gender* with numerical summary statistics below each plot to clarify the visualisation.

```{r}
#Determine amounts of males and females in the sample
length(subset(data.dropout, geslachtBin == "Male")$geslachtBin)
length(subset(data.dropout, geslachtBin == "Female")$geslachtBin)
```


```{r Descriptive 2}
#Boxplots for Big Five and Gender

#Conscientiousness
zorg <- qplot(
  x = geslachtBin,
  y = zorgvuldig,
  main = "Boxplot for Conscientiousness and Gender",
  xlab = "Gender",
  ylab = "Level of Conscientiousness",
  data = data.dropout,
  geom = "boxplot") +
    theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

zorg

by(data.dropout$zorgvuldig, data.dropout$geslachtBin, summary)

#Emotional Stability
stab <- qplot(
  x = geslachtBin,
  y = stabiel,
  main = "Boxplot for Emotional Stability and Gender",
  xlab = "Gender",
  ylab = "Level of Emotional Stability",
  data = data.dropout,
  geom = "boxplot") +
    theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

stab

by(data.dropout$stabiel, data.dropout$geslachtBin, summary)

#Openness
open <- qplot(
  x = geslachtBin,
  y = open,
  main = "Boxplot for Openness and Gender",
  xlab = "Gender",
  ylab = "Level of Openness",
  data = data.dropout,
  geom = "boxplot") +
    theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

open

by(data.dropout$open, data.dropout$geslachtBin, summary)

#Extraversion
extr <- qplot(
  x = geslachtBin,
  y = extravert,
  main = "Boxplot for Extraversion and Gender",
  xlab = "Gender",
  ylab = "Level of Extraversion",
  data = data.dropout,
  geom = "boxplot") +
    theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

extr

by(data.dropout$extravert, data.dropout$geslachtBin, summary)

#Agreeableness
altr <- qplot(
  x = geslachtBin,
  y = altrusme,
  main = "Boxplot for Agreeableness and Gender",
  xlab = "Gender",
  ylab = "Level of Agreeableness",
  data = data.dropout,
  geom = "boxplot") +
    theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

altr

by(data.dropout$altrusme, data.dropout$geslachtBin, summary)

#Uncomment to also group the plots in one window
#grid.arrange(altr, extr, open, stab, zorg, ncol= 3)
```

**Next:** Boxplots for the *Big Five* and *Foreign Origin* with numerical summaries below each plot.

```{r Descriptive 3}
#Boxplots for Big Five and foreign origin

#Conscientiousness
zorg.1 <- qplot(
  x = allochtoonBin,
  y = zorgvuldig,
  main = "Boxplot for Conscientiousness and Foreign Origin",
  xlab = "Foreign Origin",
  ylab = "Level of Conscientiousness",
  data = data.dropout,
  geom = "boxplot") +
  theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

zorg.1

by(data.dropout$zorgvuldig, data.dropout$allochtoonBin, summary)

#Emotional Stability
stab.1 <- qplot(
  x = allochtoonBin,
  y = stabiel,
  main = "Boxplot for Emotional Stability and Foreign Origin",
  xlab = "Foreign Origin",
  ylab = "Level of Emotional Stability",
  data = data.dropout,
  geom = "boxplot") +
  theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

stab.1

by(data.dropout$stabiel, data.dropout$allochtoonBin, summary)

#Openness
open.1 <- qplot(
  x = allochtoonBin,
  y = open,
  main = "Boxplot for Openness and Foreign Origin",
  xlab = "Foreign Origin",
  ylab = "Level of Openness",
  data = data.dropout,
  geom = "boxplot") +
  theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

open.1

by(data.dropout$open, data.dropout$allochtoonBin, summary)

#Extraversion
extr.1 <- qplot(
  x = allochtoonBin,
  y = extravert,
  main = "Boxplot for Extraversion and Foreign Origin",
  xlab = "Foreign Origin",
  ylab = "Level of Extraversion",
  data = data.dropout,
  geom = "boxplot") +
  theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

extr.1

by(data.dropout$extravert, data.dropout$allochtoonBin, summary)

#Agreeableness
altr.1 <- qplot(
  x = allochtoonBin,
  y = altrusme,
  main = "Boxplot for Agreeableness and Foreign Origin",
  xlab = "Foreign Origin",
  ylab = "Level of Agreeableness",
  data = data.dropout,
  geom = "boxplot") +
  theme_bw(base_family = "serif") +
    theme(axis.title.y = element_text(vjust = 1.0)) +
    theme(axis.title.x = element_text(vjust = -0.5))

altr.1

by(data.dropout$altrusme, data.dropout$allochtoonBin, summary)

#Uncomment to also group all plots in one window
#grid.arrange(arrangeGrob(zorg.1, stab.1, open.1, extr.1, altr.1, ncol = 3))
```

### Scatterplots

This section will display scatterplot for each of the *Big Five* variables on *Dropout*.

```{r Descriptive 4, echo=FALSE}
#Scatterplots Big Five on dropout
zorg <- ggplot(aes(x = zorgvuldig, y = vrv_1), 
       data = data.dropout) + 
  geom_jitter(aes(colour = allochtoonBin, 
                  shape = factor(geslachtBin), 
                  size = 0.75), 
              position = position_jitter(width = 0.3, 
                                         height = 0.3), 
                                         alpha = 0.4) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Scale of Conscientiousness", 
    y = "Dropout (Dichotomous)", 
    title = "Scatterplot Conscientiousness on Dropout") +
  theme_bw(base_family = "serif") +
  guides(shape = guide_legend("Gender"), 
         size = FALSE, 
         colour = guide_legend("Foreign Origin")) +
  geom_vline(xintercept = 1.5, alpha = 0.3) +
  geom_vline(xintercept = 2.5, alpha = 0.3) +
  geom_vline(xintercept = 3.5, alpha = 0.3) +
  geom_vline(xintercept = 4.5, alpha = 0.3) +
  geom_hline(yintercept = 0.5, alpha = 0.3)

zorg

stab <- ggplot(aes(x = stabiel, y = vrv_1), 
               data = data.dropout) + 
  geom_jitter(aes(colour = allochtoonBin, 
                  shape = factor(geslachtBin), 
                  size = 0.75), 
              position = position_jitter(width = 0.3, 
                                         height = 0.3), 
                                         alpha = 0.4) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Scale of Emotional Stability", 
    y = "Dropout (Dichotomous)", 
    title = "Scatterplot Emotional Stability on Dropout") +
  theme_bw(base_family = "serif") +
  guides(shape = guide_legend("Gender"), 
         size = FALSE, 
         colour = guide_legend("Foreign Origin")) +
  geom_vline(xintercept = 1.5, alpha = 0.3) +
  geom_vline(xintercept = 2.5, alpha = 0.3) +
  geom_vline(xintercept = 3.5, alpha = 0.3) +
  geom_vline(xintercept = 4.5, alpha = 0.3) +
  geom_hline(yintercept = 0.5, alpha = 0.3)

stab

open <- ggplot(aes(x = open, y = vrv_1), 
               data = data.dropout) + 
  geom_jitter(aes(colour = allochtoonBin, 
                  shape = factor(geslachtBin), 
                  size = 0.75), 
              position = position_jitter(width = 0.3, 
                                         height = 0.3), 
                                         alpha = 0.4) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Scale of Openness", 
    y = "Dropout (Dichotomous)", 
    title = "Scatterplot Openness on Dropout") +
  theme_bw(base_family = "serif") +
  guides(shape = guide_legend("Gender"), 
         size = FALSE, 
         colour = guide_legend("Foreign Origin")) +
  geom_vline(xintercept = 1.5, alpha = 0.3) +
  geom_vline(xintercept = 2.5, alpha = 0.3) +
  geom_vline(xintercept = 3.5, alpha = 0.3) +
  geom_vline(xintercept = 4.5, alpha = 0.3) +
  geom_hline(yintercept = 0.5, alpha = 0.3)

open

extr <- ggplot(aes(x = extravert, y = vrv_1), 
               data = data.dropout) + 
  geom_jitter(aes(colour = allochtoonBin, 
                  shape = factor(geslachtBin), 
                  size = 0.75), 
              position = position_jitter(width = 0.3, 
                                         height = 0.3), 
                                         alpha = 0.4) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Scale of Extraversion", 
    y = "Dropout (Dichotomous)", 
    title = "Scatterplot Extraversion on Dropout") +
  theme_bw(base_family = "serif") +
  guides(shape = guide_legend("Gender"), 
         size = FALSE, 
         colour = guide_legend("Foreign Origin")) +
  geom_vline(xintercept = 1.5, alpha = 0.3) +
  geom_vline(xintercept = 2.5, alpha = 0.3) +
  geom_vline(xintercept = 3.5, alpha = 0.3) +
  geom_vline(xintercept = 4.5, alpha = 0.3) +
  geom_hline(yintercept = 0.5, alpha = 0.3)

extr

altr <- ggplot(aes(x = altrusme, y = vrv_1), 
               data = data.dropout) + 
  geom_jitter(aes(colour = allochtoonBin, 
                  shape = factor(geslachtBin), 
                  size = 0.75), 
              position = position_jitter(width = 0.3, 
                                         height = 0.3), 
                                         alpha = 0.4) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Scale of Agreeableness", 
    y = "Dropout (Dichotomous)", 
    title = "Scatterplot Agreeableness on Dropout") +
  theme_bw(base_family = "serif") +
  guides(shape = guide_legend("Gender"), 
         size = FALSE, 
         colour = guide_legend("Foreign Origin")) +
  geom_vline(xintercept = 1.5, alpha = 0.3) +
  geom_vline(xintercept = 2.5, alpha = 0.3) +
  geom_vline(xintercept = 3.5, alpha = 0.3) +
  geom_vline(xintercept = 4.5, alpha = 0.3) +
  geom_hline(yintercept = 0.5, alpha = 0.3)

altr

#Uncomment to also display all plots in one window.
#grid.arrange(zorg, stab, open, extr, altr, ncol = 2)
```

In this section the distribution of dropouts among *Gender* and *Foreign Origin* will be investigated.

```{r}
#Scatterplot distribution gender and foreign origin among dropout
dropout.control <- ggplot(aes(x = geslachtBin, y = vrv_1), 
               data = data.dropout) + 
  geom_jitter(aes(colour = allochtoonBin, 
                  size = 0.75), 
              position = position_jitter(width = 0.2, 
                                         height = 0.2), 
                                         alpha = 1/3) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Gender", 
    y = "Dropout (Dichotomous)", 
    title = "Distribution of Gender and Foreign Origin among Dropout") +
  theme_bw(base_family = "serif") +
  guides(colour = guide_legend("Foreign Origin"),
         size = FALSE) +
  geom_hline(yintercept = 0.5, alpha = 0.3)

dropout.control

#Dropouts by male and female
by(data.dropout$vrv_1, data.dropout$geslachtBin, length)

#Dropouts by foreign origin and native origin
by(data.dropout$vrv_1, data.dropout$allochtoonBin, length)

#Amount of people of foreign origin in data
length(subset(data.dropout, allochtoonBin == "Yes")$allochtoonBin)

#Amount of people of native origin in data
length(subset(data.dropout, allochtoonBin == "No")$allochtoonBin)

#Create variables for either foreign origin is true or not
f1 <- subset(data.dropout, allochtoonBin == "Yes")
f0 <- subset(data.dropout, allochtoonBin == "No")

#Scatterplot dropout among foreign origin
dropout.f1 <- ggplot(aes(x = geslachtBin, y = vrv_1), 
               data = f1) + 
  geom_jitter(position = position_jitter(width = 0.095, 
                                         height = 0.095), 
                                         alpha = 0.4) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Gender", 
    y = "Dropout (Dichotomous)", 
    title = "Gender and Dropout among Foreign Origin") +
  theme_bw(base_family = "serif")

dropout.f1

#Scatterplot dropout among native origin
dropout.f0 <- ggplot(aes(x = geslachtBin, y = vrv_1), 
               data = f0) + 
  geom_jitter(position = position_jitter(width = 0.095, 
                                         height = 0.095), 
                                         alpha = 0.4) +
  scale_y_continuous(breaks = c(0, 1)) +
  labs(
    x = "Gender", 
    y = "Dropout (Dichotomous)", 
    title = "Gender and Dropout among Native Origin") +
  theme_bw(base_family = "serif")

dropout.f0

#Dropouts by male and female among foreign origin
by(f1$vrv_1, f1$geslachtBin, length)

#Dropouts by male and female among native origin
by(f0$vrv_1, f0$geslachtBin, length)

#DROPOUT FOREIGN ORIGIN
#Dropout YES foreign origin and male
length(subset(f1, vrv_1 == 1 & geslachtBin == "Male")$allochtoonBin)

#Dropout YES foreign origin and female
length(subset(f1, vrv_1 == 1 & geslachtBin == "Female")$allochtoonBin)

#Dropout NO foreign origin and male
length(subset(f1, vrv_1 == 0 & geslachtBin == "Male")$allochtoonBin)

#Dropout NO foreign origin and female
length(subset(f1, vrv_1 == 0 & geslachtBin == "Female")$allochtoonBin)

#DROPOUT NATIVE ORIGIN
#Generate native YES
n1 <- subset(data.dropout, allochtoonBin == "No")
#Generate native NO
n0 <- subset(data.dropout, allochtoonBin == "Yes")

#Dropout YES native origin and male
length(subset(n1, vrv_1 == 1 & geslachtBin == "Male")$allochtoonBin)

#Dropout YES native origin and female
length(subset(n1, vrv_1 == 1 & geslachtBin == "Female")$allochtoonBin)

#Dropout NO native origin and male
length(subset(n1, vrv_1 == 0 & geslachtBin == "Male")$allochtoonBin)

#Dropout NO native origin and female
length(subset(n1, vrv_1 == 0 & geslachtBin == "Female")$allochtoonBin)
```

### Correlation Matrix

```{r, eval=FALSE}
CorMatrix <- cor(data.dropout)
corrplot(CorMatrix, method = "circle", type = "lower", order ="AOE")
```

